Data Management \\ and LSST Special Programs

### 2017-09-01 ms.tex
finished with case studies; added K-T feedback

\setDocAbstract{\textbf{WORKING DRAFT (contains rough notes and partial thoughts) } but intended to be an overview of any and all potential requirements on data management (DM) from Special Programs such as the deep drilling fields (DDF) or mini-surveys. We also summarize the plans for how the Special Programs data will be integrated into the wide-fast-deep (WFD) Main Survey data products(Levels 1 and 2). It is the intent that this document evolve to become an internal catalog of change requests to the DM Level 1 and 2 pipelines and products, and a community resource for future Special Program white papers and preparations for Level 3 pipelines.}

@@

The intent of LSST DM for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}):

\vspace{-20pt}
\begin{quote}
LSST will not write unique algorithms for processing Special Programs data or reprocessing Main Survey data, but LSST will reconfigure the pipelines and generate imaging and catalog products for Special Programs data whenever possible, and LSST will make its codes accessible to the science community and commit $\sim$10\% of its computing resources toward enabling Level 3 analysis and data product creation, including user-driven Special Programs processing.
\end{quote}

\noindent The purpose of this study is to ensure that the requirements and plans of DM, as written down in all major DM documentation, accurately reflects this intent. In particular, this study is designed to review: \\
(1) the processing and products that are required to enable science for Special Programs, \\
(2) the current plans for DM processing and products with respect to Special Programs, \\
(3) whether the DM plans meet the expected needs of Special Programs, and \\
(4) any necessary changes to be written into the requirements, designs, and plans.

\noindent The concerns raised by this study fall into three main categories: \\
(1) Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be. \\
(2) Ensuring that DM's intent to process Special Programs data with reconfigured pipelines and provide separate data products that meet the science goals of each program is accurately reflected in all documents, as it is in the DPDD. \\
(3) Clarifying the computational sizing plans for processing and storage of Special Programs data and whether it is this separate from the allocation for end-user Level 3 analysis.

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\section{The Potential Diversity of Special Programs Data} \label{ssec:data}

@@

\noindent {\bf Resources:} \\
$\bullet$ ``LSST: from Science Drivers to Reference Design and Anticipated Data Products" Ivezi\'{c} et al. (2008), \cite{2008arXiv0805.2366I} \\
$\bullet$ The LSST Science Requirements Document (SRD), LPM-17, \cite{LPM-17} \\
$\bullet$ LSST Deep Drilling white papers from 2011: \url{https://project.lsst.org/content/whitepapers32012} \\
$\bullet$ ``General Review of the Proposed DDF and MS", LSST AHM Aug 2016 presentation by Niel Brandt \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf} \\
$\bullet$ ``Simulations, Metrics and Merit Function for Mini-Surveys and DDF", LSST AHM Aug 2016 presentation by Stephen Ridgway \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf} \\
$\bullet$ ``LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \cite{2013arXiv1304.3455G} \url{https://arxiv.org/abs/1304.3455} \\
$\bullet$ Chapter 10 ``Special Surveys" of the Observing Strategy White Paper \cite{2017arXiv170804058L}

@@

\textbf{A Nominal DDF Observing Strategy -- } Ivezi\'{c} et al. (2008, \cite{2008arXiv0805.2366I}), Section 3.2.1 ``Mini-Surveys": describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters in one hour per night, once every two nights, for four months. Each observation would have a limit of $r\sim24.5$; a one-hour nightly stack would have a limit of $r\sim26.5$; and and assuming a $60\%$ completion rate (weather), the four-month $\sim40$ hours stacked together with the $\sim180$ main survey visits would yield a limit of $r\sim28$.

\medskip
\noindent \textbf{Solar System Objects (SSO)}\\
$\bullet$ North Ecliptic Spur: observations in this area of the sky will yield more $\geq140$ m near-earth objects (NEOs) for the final LSST sample (Brandt talk). \\
$\bullet$ Mini-Moons: finding and studying temporarily captured satellites of the Earth. (Section 10.2, \cite{2017arXiv170804058L}) \\
$\bullet$ Sweetspot: use twilight fields to find NEOs in Earth-like orbits (never in opposition fields). (Section 10.2, \cite{2017arXiv170804058L}) 
$\bullet$ Meter-Sized Impactors: finding and tracking meter-sized impactors $<2$ weeks before impact.  (Section 10.2, \cite{2017arXiv170804058L}) \\
$\bullet$ Detecting faint SSO down to $r\sim27$ would be done by applying shift-and-stack (SAS) processing to a mini-survey data of standard visit images (faint SSO include: Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) \citep{BeckerWP}. Regarding SAS, \cite{BeckerWP} says that ``the multi-fit algorithm ... naturally provides a base infrastructure for this process. In particular, the marshaling of the pixels to attempt a given photometric measurement is non-trivial when tens of thousands of images are required. However, the multi-fit middleware is required to do exactly this, so we expect that this issue will be resolved by the time SAS is needed." \\
$\bullet \bullet \bullet$ {\bf Summary.} Most of these science goals can be achieved with standard visit images obtained in a normal survey pattern taken during the night. The exception is a Sweetspot survey, which occurs during twilight and thus may require shorter exposures. Many of these science goals will also be possible by using the products of Moving Object Processing System (MOPS), which runs on the Level 1 {\tt DIASource} catalogs updated each night. The exception is finding faint SSOs through shift-and-stack processing; SAS is not a capability of the DM system, cannot be done solely by reconfiguring DM pipelines, and is therefore considered Level 3. See Sections \ref{ssec:SPCS_SAS} and \ref{ssec:SPCS_MM} for more detailed DM processing case studies.

\noindent \textbf{Stars in the Milky Way and Magellanic Clouds} \\
$\bullet$ When surveying the Galactic Plane, obtaining all the images over a short time span would mitigate proper motion loss (and increase flare detection rates) and help identify useful stellar populations. When targeting Galactic plane regions and/or open clusters, typically 1 to 3 filters are needed for detections of e.g., faint stars already detected in $z$ and $y$ but needing $i$ to distinguish from red galaxies \cite{DhitalWP}. \\
$\bullet$ A nominal deep-drilling observing strategy over the full L/SMC galaxies will characterize stellar variability to $M_V<6.5$ on timescales from 15s to 3d \cite{SzkodyWP}. For this, special co-adds may be required, e.g., {\it "to reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"}. \citep{SzkodyWP}. \\
$\bullet$ A Twilight Short Exposure survey would enable bright stars, which have longer monitoring baselines due to historical observations, to be put on the same photometric system as the deeper LSST WFD main survey catalog. \\
$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it ``custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}. \\
$\bullet \bullet \bullet$ {\bf Summary.} Many of these science goals can be achieved with standard visit images, but with alternative observing strategies such as limiting the number of filters or concentrating the depth build-up into shorter timeframes; the exception is short-exposure surveys to obtain bright stars. Similarly, most of these science goals can be met by reconfiguring the planned DM pipelines to detect sources, measure proper motion, and create deep stacks. The exception might be in processing exposures as short as 0.1 seconds, and attempting to measure flux of a saturated stars. See Sections \ref{ssec:SPCS_GPVSEx} and \ref{ssec:SPCS_M67} for more detailed DM processing case studies.

@@

\noindent \textbf{Supernovae} \\
$\bullet$ The nominal DDF plan described in \cite{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$ out of standard visit images, would extend the SN sample to $z\sim1.2$ and provide more densely sampled light curves for cosmological analyses. The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$. \cite{KesslerWP} \\
$\bullet$ High-cadence observations of DDF would be the only way to detect fast transients, particularly extragalactic novae, some tidal disruption events, optical counterparts to gamma-ray bursts, and peculiar SNe \cite{2014ApJ...794...23D}. \\
$\bullet$ Generating the best-possible individual SN light curves for cosmological analyses requires building special, deep-as-possible, SN-free host galaxy images and using them as a template. This will also be necessary for studying SNe that appear in the template image; i.e., that last $>1000$ days. These are mostly Type IIn, probably explosions of massive stars into dense circumstellar material, which are not used for cosmology but rather to study late-stage stellar evolution and mass loss. SN-free images will also be needed to measure correlated properties for cosmology and to do host-galaxy science. The latter, specifically the ``characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \cite{FergusonWP}. \\
$\bullet$ \textit{MLG - I'm not sure this has been yet proposed, but short-exposure observations to include nearby SNe on the same photometric system could be useful?} \\
$\bullet \bullet \bullet$ {\bf Summary.} All of these science goals appear possible with standard visit images (with the exception of an as-yet unproposed target-of-opportunity survey to obtain photometry for nearby, saturated SNe). Additionally, all are accessible with reconfigured DM pipelines to stack and difference the data. In particular, the Level 2 DRP codes to create ``transient-free CoAdds" will be suitable for generating the SN-free templates for DDF, as they will do for the Main Survey images. See also Section \ref{ssec:SPCS_SNDDF} for a DM processing case study to find SNe in a DDF.

\noindent \textbf{Galaxies} \\
$\bullet$ Use the additional depth of a DDF to build a large collection of low-$\mu$ objects. \cite{FergusonWP} mentions ``identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and ``characterization of low-surface-brightness extended features around both nearby and distant galaxies". \\
$\bullet$ The DDF stacks can also be used to characterize of high-$z$ clusters, although this ability might depend on deblending extended objects. \\
$\bullet$ The DDF observations, when combined with the WFD, allow for AGN monitoring on a variety of timescales in well-characterized galaxies \citep{FergusonWP,GawiserWP}. \\
$\bullet \bullet \bullet$ {\bf Summary.} As with the SN science goals, these use standard visit images and reconfigured DM pipelines to make deep CoAdds and extract sources. In addition, it seems likely that Level 3 algorithms that are optimized to detect and characterize particular types of faint extended sources will be needed, but these are beyond the scope of DM.

\noindent \textbf{Weak Lensing} \\
$\bullet$ The deeper imaging from DDFs can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch) \\
$\bullet$ {\it Jim Bosch -- ``Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} ({\tt \Large{Community}} forum) \\
$\bullet$ {\it Jim Bosch -- ``Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline. We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} ({\tt \Large{Community}} forum)  \\
$\bullet$ {\it MLG side note -- Photo-$z$ are very important to weak lensing \citep{MaWP} and so perhaps the implemented method should be chosen with weak lensing science prioritized.} \\
$\bullet \bullet \bullet$ {\bf Summary.} As with the SN and Galaxies DDF-related science goals, these use standard visit images and reconfigured DM pipelines can be used to make deep CoAdds and extract sources, as Jim notes.

@@

\textbf{Initial community suggestions regarding the call for white papers.} \\
$\bullet$ The call should describe the reconsideration timescale for evaluating Special Programs' progress and the power of the Project/SAC to terminate/initiate a Special Program. \\
$\bullet$ The call should be clear about whether a single white paper can be focused on a single science goal or a single observing strategy or, for example, to what extent they should incorporate all possible science for a given proposed set of observations. Sample outlines for white papers that are driven by a single science goal, or by a novel survey strategy, might help the proposers. \\
$\bullet$ Regarding the prior point, being clear about how multiple white papers might be combined into an ``eigenset" of mini-surveys at the recommendation of the SAC will probably help. \\
$\bullet$ Proposers will be asked to include "Data Management Considerations for Special Programs" into their white papers, and so they will also need a readable description of DM's plans and boundaries for Special Programs data (i.e., what this document is attempting to build).

@@

\noindent The questions opened up on that forum are: \\
(1) What additional processing beyond that currently planned by the DM team (alerting relative to an annually created template) would greatly enhance the DDF science goals? \\
(2) Are there DDF or Mini-Surveys specific aspects of the Level 3 system that would add significant value if provided? ``Level 3" is the LSST-provided capability that enables non-DM, user-driven, processing of LSST data at the LSST Archive center (or remotely). \\
(3) Are there aspects of the Science User Interface and Tools (SUIT) that need to be developed in order to enhance the usefulness of DDF data products. \\
(4) To what degree should the DDF or Mini-Survey imaging could/should be incorporated into the main survey's deep stacks and associated data products (as opposed to being processed as separate data products)?

@@

It might be helpful for the community to know, at the time of the call for new white papers, if certain non-standard visit images cannot processed by the DM pipelines. Exposures which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in difference imaging and the correction for differential chromatic refraction, and/or photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD. We consider shorter and longer exposure in turn.

(1) Shorter Exposures. The minimum supported exposure time is currently specified as 1 second (stretch goal 0.1 seconds), but this is a camera constraint. The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint. Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes. A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap). However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale). As a simple demonstration, MLG has made Figure \ref{fig:expt} to show that perhaps exposure times shorter than 2 seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for ``well-formed"). Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.

@@

Our review of the technical boundaries and previously proposed Special Programs leads us to conclude that the potential diversity of data is mainly limited to exposures that are significantly shorter than the WFD main survey standard visit exposures of $2\times15$ seconds, and exposures that are obtained with a bright sky background during twilight.

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\section{The Current DM Plans for Special Programs Data Processing} \label{sec:dmplans}

@@

%\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C1} There are currently no requirements on DM to incorporate data from Special Programs; should there be? \end{enumerate}
%\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C2} Should DM undertake an investigation to define the exposure boundaries (integration time, sky background) for non-standard visit images that can be processed by DM pipelines, such as ISR and DIA? \end{enumerate}
%\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C3} Given the double-processing of Special Programs data, is the system sized appropriately? \end{enumerate}

@@

The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data. Since MOPS takes {\tt DIASources} as input, any SP images that can be run through the Alert Pipeline can be ingested by MOPS. As discussed under ``Solar System Objects (SSO)" in Section \ref{ssec:data_science}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway.

@@

This document does contain requirements for the data that are needed to achieve the main science goals, but the version available at this time is from 7/6/2011, and some specifications are out of date (e.g., the minimum exposure time is set to 5 seconds, in contrast with OSS-REQ-0291). The SRD makes two mentions of Special Programs: that the ``LSST is well suited to conducting Deep Supernova Survey" (Section 2.3), and that 90\% of the observing time will be spent on the main survey, with the remaining 10\% used for a variety of other programs (Section 3.4). The latter we know to be an approximate estimate.

@@

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\section{Special Programs Processing Case Studies}\label{sec:SPCS}

>>>>

\setDocAbstract{The purpose of this study is to ensure that the LSST Data Management's plans for processing data from Special Programs (Deep Drilling Fields and Mini-Surveys) will meet the needs of the scientific community. To do this, we begin with a review of the broad science goals of the previously proposed Special Programs and the potential diversity of data they might generate. We also consider any technical limitations on the data from the telescope, camera, and/or DM processing capabilities, and the methods by which information could be exchanged with the scientific community on these issues (i.e., the upcoming call for white paper proposals). We then move on to summarize the current DM plans for processing Special Programs data, and highlight specific requirements or text in the existing documentation that might need additions or clarifications to reflect these plans. We also develop a series of case studies for DM's processing of Special Programs data to better illuminate any complex or hidden issues; these case studies could also serve as examples for the upcoming call for community white papers.}

@@

The intent of the LSST Data Management team for processing data from Special Programs can be summarized by the following statement, which reflects Section 6 of the Data Products Definitions Document (DPDD, LSE-163, \cite{LSE-163}):

\noindent The purpose of this study is to ensure that the requirements and plans of DM, as written down in all major DM documentation, accurately reflects this intent. In particular, this study is designed to review: (1) the processing and products that are required to enable science for Special Programs, (2) the current plans for DM processing and products with respect to Special Programs, (3) whether the DM plans meet the expected needs of Special Programs, and (4) any necessary changes to be written into the requirements, designs, and plans.

\subsection{Study Overview}\label{ssec:intro_overview}

Section \ref{sec:data} compiles a comprehensive overview of the potential diversity in data from Special Programs, compared to the Wide-Fast-Deep (WFD) main survey data: standard visit images of $2\times15$ seconds for a wide area of the nighttime sky, obtained with a dither pattern using up to five filters. We assemble the possible limitations on Special Programs observations from the capabilities of the telescope and camera, the scientific needs of as many currently proposed Special Programs as we can find, and the future mechanisms for LSST to gather more community input. We also consider whether DM might need to impose boundaries on the parameters of non-standard images that its pipelines will be able to processes.

Section \ref{sec:dmplans} reviews the current DM plans for processing Special Programs data: reconfiguring pipelines to create unique products, incorporating data into the WFD main survey products, and enabling end-user Level 3 processing. Having established the {\it intent} of DM, we step through each of the relevant LSST documents and raise {\bf "Concerns"}, which highlight specific requirements or text in these documents that might need additions or clarifications to properly reflect DM's plans for processing Special Programs data. These {\bf Concerns} are not prioritized by urgency or importance, but instead represent more of a complete set of possible issues. {\bf The Concerns raised by this study fall into two main categories:}
\begin{enumerate}
\item Whether DM will be able to process the full diversity of non-standard visit images, and if not, what the boundaries will be.
\item Ensuring that DM's intent to process Special Programs data with reconfigured pipelines, and to provide separate data products that meet the science goals of each program, is accurately reflected in all documents, as it is in the DPDD.
\end{enumerate}

Section \ref{sec:SPCS} contains a series of case studies for DM's processing of Special Programs data, which are based on several of the currently circulating ideas for deep drilling fields and mini-surveys. These processing case studies contribute to this work by offering an independent vantage point from which to identify a few additional {\bf Concerns} for the processing of Special Programs data, and could also serve as examples for the upcoming call for community white papers, which will request that a similar section be included in all proposals.

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\clearpage
\section{The Potential Diversity of Special Programs Data} \label{sec:data}

@@

\noindent {\bf Resources:} \\
$\bullet$ "LSST: from Science Drivers to Reference Design and Anticipated Data Products" Ivezi\'{c} et al. (2008), \cite{2008arXiv0805.2366I} \\
$\bullet$ The LSST Science Requirements Document (SRD), LPM-17, \cite{LPM-17} \\
$\bullet$ LSST Deep Drilling white papers from 2011: \url{https://project.lsst.org/content/whitepapers32012} \\
$\bullet$ "General Review of the Proposed DDF and MS", LSST AHM Aug 2016 presentation by Niel Brandt \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Brandt-DDF-MiniSurveys-01.pdf} \\
$\bullet$ "Simulations, Metrics and Merit Function for Mini-Surveys and DDF", LSST AHM Aug 2016 presentation by Stephen Ridgway \url{https://project.lsst.org/meetings/lsst2016/sites/lsst.org.meetings.lsst2016/files/Ridgway-SimulationsMetrics_1.pdf} \\
$\bullet$ "LSST's DC [Deep CoAdd] Bias Against Planets and Galactic-Plane Science" by A. Gould, \cite{2013arXiv1304.3455G} \url{https://arxiv.org/abs/1304.3455} \\
$\bullet$ Chapter 10 "Special Surveys" of the Observing Strategy White Paper \cite{2017arXiv170804058L}

@@

\textbf{A Nominal DDF Observing Strategy -- } Ivezi\'{c} et al. (2008, \cite{2008arXiv0805.2366I}), Section 3.2.1 "Mini-Surveys": describes a nominal DDF data set as $\sim50$ consecutive $15$ second exposures in each of four filters in one hour per night, once every two nights, for four months. Each observation would have a limit of $r\sim24.5$; a one-hour nightly stack would have a limit of $r\sim26.5$; and and assuming a $60\%$ completion rate (weather), the four-month $\sim40$ hours stacked together with the $\sim180$ main survey visits would yield a limit of $r\sim28$.

\medskip
\noindent \textbf{Solar System Objects (SSO)}\\
$\bullet$ North Ecliptic Spur: observations in this area of the sky will yield more $\geq140$ m near-earth objects (NEOs) for the final LSST sample (Brandt talk). \\
$\bullet$ Mini-Moons: finding and studying temporarily captured satellites of the Earth. (Section 10.2, \cite{2017arXiv170804058L}) \\
$\bullet$ Sweetspot: use twilight fields to find NEOs in Earth-like orbits (never in opposition fields). (Section 10.2, \cite{2017arXiv170804058L}) \\ 
$\bullet$ Meter-Sized Impactors: finding and tracking meter-sized impactors $<2$ weeks before impact.  (Section 10.2, \cite{2017arXiv170804058L}) \\
$\bullet$ Detecting faint SSO down to $r\sim27$ would be done by applying shift-and-stack (SAS) processing to a mini-survey data of standard visit images (faint SSO include: Trans-Neptunian Objects Trojans, asteroids, long-period comets, dwarf planets) \citep{BeckerWP}. Regarding SAS, \cite{BeckerWP} says that "the multi-fit algorithm ... naturally provides a base infrastructure for this process. In particular, the marshaling of the pixels to attempt a given photometric measurement is non-trivial when tens of thousands of images are required. However, the multi-fit middleware is required to do exactly this, so we expect that this issue will be resolved by the time SAS is needed." \\
$\bullet \bullet \bullet$ {\bf Summary.} Most of these science goals can be achieved with standard visit images obtained in a normal survey pattern taken during the night. The exception is a Sweetspot survey, which occurs during twilight and thus may require shorter exposures. Many of these science goals will also be possible by using the products of Moving Object Processing System (MOPS), which runs on the Level 1 {\tt DIASource} catalogs updated each night. The exception is finding faint SSOs through shift-and-stack processing; SAS is not a capability of the DM system, cannot be done solely by reconfiguring DM pipelines, and is therefore considered Level 3. See Sections \ref{ssec:SPCS_SAS} and \ref{ssec:SPCS_TNO} for more detailed DM processing case studies.

\noindent \textbf{Stars in the Milky Way and Magellanic Clouds} \\
$\bullet$ When surveying the Galactic Plane, obtaining all the images over a short time span would mitigate proper motion loss (and increase flare detection rates) and help identify useful stellar populations. When targeting Galactic plane regions and/or open clusters, typically 1 to 3 filters are needed for detections of e.g., faint stars already detected in $z$ and $y$ but needing $i$ to distinguish from red galaxies \cite{DhitalWP}. \\
$\bullet$ A nominal deep-drilling observing strategy over the full L/SMC galaxies will characterize stellar variability to $M_V<6.5$ on timescales from 15s to 3d \cite{SzkodyWP}. For this, special co-adds may be required, e.g., {\it "to reach variability levels of 0.1 to 0.005 mag will require co-adds depending on the timescale of the particular variables"}. \citep{SzkodyWP}. \\
$\bullet$ A Twilight Short Exposure survey would enable bright stars, which have longer monitoring baselines due to historical observations, to be put on the same photometric system as the deeper LSST WFD main survey catalog. \\
$\bullet$ A short-exposure survey of M67, Chapter 10.4 of \cite{2017arXiv170804058L}, suggests using the stretch goal of 0.1 second exposures or, if that is not possible, {\it "custom pixel masks to accurately perform photometry on stars as much as 6 magnitudes brighter than the saturation level"}. \\
$\bullet \bullet \bullet$ {\bf Summary.} Many of these science goals can be achieved with standard visit images, but with alternative observing strategies such as limiting the number of filters or concentrating the depth build-up into shorter timeframes; the exception is short-exposure surveys to obtain bright stars. Similarly, most of these science goals can be met by reconfiguring the planned DM pipelines to detect sources, measure proper motion, and create deep stacks. The exception might be in processing exposures as short as 0.1 seconds, and attempting to measure flux of a saturated stars. See Section \ref{ssec:SPCS_GPVSEx} for more detailed DM processing case studies.

@@

\noindent \textbf{Supernovae} \\
$\bullet$ The nominal DDF plan described in \cite{2008arXiv0805.2366I}, which builds nightly stacks with a limit of $r\sim26.5$ out of standard visit images, would extend the SN sample to $z\sim1.2$ and provide more densely sampled light curves for cosmological analyses. The optimal exposure time distribution might be 6, 5, 10, 10, 9, 10 in $ugrizy$. \cite{KesslerWP} \\
$\bullet$ High-cadence observations of DDF would be the only way to detect fast transients, particularly extragalactic novae, some tidal disruption events, optical counterparts to gamma-ray bursts, and peculiar SNe \cite{2014ApJ...794...23D}. \\
$\bullet$ Generating the best-possible individual SN light curves for cosmological analyses requires building special, deep-as-possible, SN-free host galaxy images and using them as a template. This will also be necessary for studying SNe that appear in the template image; i.e., that last $>1000$ days. These are mostly Type IIn, probably explosions of massive stars into dense circumstellar material, which are not used for cosmology but rather to study late-stage stellar evolution and mass loss. SN-free images will also be needed to measure correlated properties for cosmology and to do host-galaxy science. The latter, specifically the "characterization of ultra-faint SN host galaxies", is also mentioned in the Galaxies DDF WP \cite{FergusonWP}. \\
$\bullet$ \textit{MLG - I'm not sure this has been yet proposed, but short-exposure observations to include nearby SNe on the same photometric system could be useful?} \\
$\bullet \bullet \bullet$ {\bf Summary.} All of these science goals appear possible with standard visit images (with the exception of an as-yet unproposed target-of-opportunity survey to obtain photometry for nearby, saturated SNe). Additionally, all are accessible with reconfigured DM pipelines to stack and difference the data. In particular, the Level 2 DRP codes to create "transient-free CoAdds" will be suitable for generating the SN-free templates for DDF, as they will do for the Main Survey images. See also Section \ref{ssec:SPCS_SNDDF} for a DM processing case study to find SNe in a DDF.

\noindent \textbf{Galaxies} \\
$\bullet$ Use the additional depth of a DDF to build a large collection of low-$\mu$ objects. \cite{FergusonWP} mentions "identification of nearby isolated low-redshift dwarf galaxies via surface-brightness fluctuations" and "characterization of low-surface-brightness extended features around both nearby and distant galaxies". \\
$\bullet$ The DDF stacks can also be used to characterize of high-$z$ clusters, although this ability might depend on deblending extended objects. \\
$\bullet$ The DDF observations, when combined with the WFD, allow for AGN monitoring on a variety of timescales in well-characterized galaxies \citep{FergusonWP,GawiserWP}. \\
$\bullet \bullet \bullet$ {\bf Summary.} As with the SN science goals, these use standard visit images and reconfigured DM pipelines to make deep CoAdds and extract sources. In addition, it seems likely that Level 3 algorithms that are optimized to detect and characterize particular types of faint extended sources will be needed, but these are beyond the scope of DM.

\noindent \textbf{Weak Lensing} \\
$\bullet$ The deeper imaging from DDFs can help with shear systematics and the effects of magnification in the analysis of WFD data (community forum, Jim Bosch) \\
$\bullet$ {\it Jim Bosch -- "Will need to process at least some deep drilling fields (high-latitude ones) in the same way we process a full data release production before running the full data release production, so we can use the results to build priors and/or calibrate shear estimates on the wide survey"} ({\tt \Large{Community}} forum) \\
$\bullet$ {\it Jim Bosch -- "Will need to process various wide-depth subsets of some deep drilling fields (again, high-latitude ones) using the regular DRP pipeline. We'll definitely want best-seeing, worst-seeing, and probably a couple of independent typical-seeing subsets, but there may be other ways we'd want to subdivide as well."} ({\tt \Large{Community}} forum)  \\
$\bullet$ {\it MLG side note -- Photo-$z$ are very important to weak lensing \citep{MaWP} and so perhaps the implemented method should be chosen with weak lensing science prioritized.} \\
$\bullet \bullet \bullet$ {\bf Summary.} As with the SN and Galaxies DDF-related science goals, these use standard visit images and reconfigured DM pipelines can be used to make deep CoAdds and extract sources, as Jim notes.

@@

\textbf{Initial community suggestions regarding the call for white papers.} \\
$\bullet$ The call should describe the reconsideration timescale for evaluating Special Programs' progress and the power of the Project/SAC to terminate/initiate a Special Program. \\
$\bullet$ The call should be clear about whether a single white paper can be focused on a single science goal or a single observing strategy or, for example, to what extent they should incorporate all possible science for a given proposed set of observations. Sample outlines for white papers that are driven by a single science goal, or by a novel survey strategy, might help the proposers. \\
$\bullet$ Regarding the prior point, being clear about how multiple white papers might be combined into an "eigenset" of mini-surveys at the recommendation of the SAC will probably help. \\
$\bullet$ Proposers will be asked to include "Data Management Considerations for Special Programs" into their white papers, and so they will also need a readable description of DM's plans and boundaries for Special Programs data (i.e., what this document is attempting to build).

@@

\noindent The questions opened up on that forum are: \\
(1) What additional processing beyond that currently planned by the DM team (alerting relative to an annually created template) would greatly enhance the DDF science goals? \\
(2) Are there DDF or Mini-Surveys specific aspects of the Level 3 system that would add significant value if provided? "Level 3" is the LSST-provided capability that enables non-DM, user-driven, processing of LSST data at the LSST Archive center (or remotely). \\
(3) Are there aspects of the Science User Interface and Tools (SUIT) that need to be developed in order to enhance the usefulness of DDF data products. \\
(4) To what degree should the DDF or Mini-Survey imaging could/should be incorporated into the main survey's deep stacks and associated data products (as opposed to being processed as separate data products)?

@@

It might be helpful for the community to know, at the time of the call for new white papers, if certain non-standard visit images cannot processed by the DM pipelines, such as those with different exposure time, dither pattern, or sky background. Alternatively, it might make more sense that DM's processing boundaries apply not to the exposure acquisition parameters, but the image quality parameters such as number of stars with sufficient flux for calibration or the source area density (crowding). The boundaries of this parameter space might not be fully constrained until the final performance of DM algorithms as described in the Data Management Applications Design (DMAD, LDM-151, \cite{LDM-151}) document are characterized (see also Section \ref{sssec:dmplans_review_dmad}).

Regarding exposure time, images which deviate significantly from the $15$ second duration for the WFD main survey may encounter issues in the instrument signature removal routine, in difference imaging and the correction for differential chromatic refraction, and/or photometric and astrometric calibrations due to a differently sampled set of standard stars per CCD. We consider shorter and longer exposure in turn.

(1) Shorter Exposures. The minimum supported exposure time is currently specified as 1 second (stretch goal 0.1 seconds), but this is a camera constraint. The minimum exposure time for an image to be successfully reduced with Instrument Signature Removal (ISR) will be a DM constraint. Assuming that 1 second exposure can be reduced and calibrated, its detected point sources will span a dynamic range of $r \approx 12.9$ -- $21.0$ magnitudes. A template image built on $15$ second exposures will saturate at $r \approx 15.8$, but this still leaves stars between $15.8$--$21.0$ magnitudes to be used in the PSF-matching (and all other filters have a similarly large overlap). However, in order for an image to be successfully PSF-matched to the template, the PSF must be well formed (i.e., no speckle pattern), and have a spatial variance that the pipeline is capable of modeling (i.e., be smoothly varying on some minimal scale). As a simple demonstration, MLG has made Figure \ref{fig:expt} to show that perhaps exposure times shorter than 2 seconds do not have a well-formed PSF (using the centroid of a 2D Gaussian fit as a proxy for "well-formed"). Another concern is that the Level 1 pipeline would not be able to deliver the required $60$ second latency if the camera produces more than 1 image in 30 seconds, which would disqualify a contribution to the Alert Stream.

@@

Our review of the technical boundaries and currently tabled Special Programs leads us to conclude that the potential diversity of data is mainly limited to exposures that are significantly shorter than the WFD main survey standard visit exposures of $2\times15$ seconds, exposures that are obtained with a bright sky background during twilight, and images of crowded fields.

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\clearpage
\section{The Current DM Plans for Special Programs Data Processing} \label{sec:dmplans}

@@

@@

The WFD main survey observations will be optimized to meet the mandated goals for moving object discoveries, so it is unclear whether the science goals of MOPS can be much improved by incorporating Special Programs data. Since MOPS takes {\tt DIASources} as input, any SP images that can be run through the Alert Pipeline can be ingested by MOPS. As discussed under "Solar System Objects (SSO)" in Section \ref{ssec:data_science}, most of the Special Programs data associated with SSO science will obtain standard visit images anyway.

@@

This document does contain requirements for the data that are needed to achieve the main science goals, but the version available at this time is from 7/6/2011, and some specifications are out of date (e.g., the minimum exposure time is set to 5 seconds, in contrast with OSS-REQ-0291). The SRD makes two mentions of Special Programs: that the "LSST is well suited to conducting Deep Supernova Survey" (SRD, Section 2.3), and that 90\% of the observing time will be spent on the main survey, with the remaining 10\% used for a variety of other programs (SRD, Section 3.4). The latter we know to be an approximate estimate.

@@

% % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % % 
\clearpage
\section{Special Programs Processing Case Studies}\label{sec:SPCS}
